---
title: "Project 1 Fish McDowell"
format: pdf
editor: visual
---

```{r}
#| include: false
library(tidyverse)
```


Data is everywhere. It's power is immeasurable with finding patterns, modeling relationships, and driving decisions. In order to be able to do that, data must be handled appropriately.  In this report, we will go through the motions of loading in and preprocessing some data so that it's true power can be used as discussed above.


## Question 1: Selecting Columns

First, we will load in the appropriate data set and select only `area_name`, `STCOU`, and any columns that end with the letter D.

```{r}
sec1 <- read_csv("./data/EDU01a.csv", col_names = TRUE)

sec1_new <- sec1 |>
  select(area_name = Area_name,
         STCOU,
         ends_with("D"))

head(sec1_new, n = 5)
```

The selected columns look to be what we hoped.

## Question 2: Long Formatted Data

Next, we will convert this data into long format with only one row per enrollment value for that area name. We will put the column names into a separate new variable to keep that information.


```{r}
sec1_long <- sec1_new |>
  pivot_longer(cols = 3:12,
               names_to = "survey_type",
               values_to = "vals")

head(sec1_long, n = 5)
```

This looks to match the pivot that we hoped to make.



2. Convert the data into long format where each row has only one enrollment value for that Area_name.
Display the first 5 rows of your new data set to show that you created this correctly.

